\begin{Exc}{(A5.16.v0, 5 points):}
Consider an alternative validity condition for binary consensus: There must
be at least one admissible execution with decision value 0, and there
must be at least one admissible execution with decision value 1.
Show that there is no wait-free algorithm for this modified consensus
problem in single-writer shared memory asynchronous systems.

\normalfont
Hint: Modify the proof of the existence of a bivalent initial
configuration for ordinary consensus in the textbook.
\end{Exc}
\begin{Exc}{(L6.1.v0, 4 points):}
It is known that one can solve synchronous consensus
in the presence of $f=1$ Byzantine faulty processors
in a fully connected system of $n=4$ processors. Using an easy
impossibility proof, show that this is no longer possible when
two processors cannot communicate directly with each other.

\normalfont
Try find a topology
consisting of correct processors only, which appears,
viewed from the local perspective of every processor, like
a 4-processor system with at most 1 Byzantine one where one
link is missing.
\end{Exc}
\begin{Exc}{(S.5.4.v0, 7 points):}
Consider a lockstep synchronous system, where processors never fail but
messages can be lost arbitrarily. Prove that it is impossible to solve
binary consensus with the alternative validity property
\begin{itemize}
\item If all processors start with the same
input value, then every processor~$p_i$ computes
\begin{itemize}
%\itemsep-2pt
\item $y_i=1$ if $\forall j: x_j=1$ and no message got
lost in the entire execution,
\item $y_i=0$ if $\forall j: x_j=0$,
\end{itemize}
\end{itemize}
in a system of $n=2$ processors. Does this impossibility also
hold for the following variants of consensus:
\begin{enumerate}
\item[(a)] Ordinary consensus, i.e., agreement, termination and
the standard validity property from the textbook?
\item[(b)] Consensus with unanimous termination, i.e., agreement and
validity from the textbook, and the requirement that
\begin{itemize}
\item if some processor decides, all processors eventually decide, plus
\item if no message is lost, all processors eventually decide?
\end{itemize}
\end{enumerate}

\normalfont
Hint: Start out from a failure-free execution where $x_0=x_1=1$.
Show that you can gradually drop all messages without
changing the decision value. Remember that any processor's
decision is irreversible.
\end{Exc}
\begin{Exc}{(S.5.14.v0, 6 points):}
Consider synchronous consensus with at most $f$ crash failures in not
fully connected communication graphs $G$.
\begin{enumerate}
\item[(1)] Using a partitioning argument, prove that there is no solution for this problem if $G$
is not $f+1$-connected.
\item[(2)] Extend the proof of the termination time lower bound of $f+1$ rounds
for fully connected communication graphs: Prove that every algorithm ${\cal A}$
(that may know $G$) for $n\geq f+2$ processors needs at least $f+Radius(G)$ rounds
for termination.

Research problem (optional): You may win extra points by either improving this
bound (I conjecture that the actual bound is $f+Diam(G)$) or else
by giving an algorithm and a proof that it indeed terminates in
$f+Radius(G)$.
\end{enumerate}
\end{Exc}
