\newcommand{\off}[1]{\text{off}_{#1}}
\newcommand{\rev}[1]{\text{rev}(#1)}

\begin{Exc}{(A3.9.v0, 4 points):}
Consider the ring $R_n^{rev}$, where processor $p_i$ has UID
$rev(i)$, $0\leq i \leq n-1$ for $n=2^k$. Prove that all
consecutive segments of size $2^\ell$, $0<\ell < k$, in $R_n^{rev}$
are order-equivalent.
\end{Exc}

\begin{lemma} \label{lemma:msb_order}
Let $x, y$ be $k$-bit integers with non-identical $\ell$ most significant bits
which we call $x_{MSB}, y_{MSB}$, respectively. Then the ordering relation of $x, y$ is
completely determined by $x_{MSB}, y_{MSB}$.
\end{lemma}

\begin{proof}
Let $i$ be the position of the highest differing bit between $x, y$, and let
its value be denoted $x_i, y_i$. Note that
$k - \ell \leq i < k$ since $x_{MSB}, y_{MSB}$ are non-identical. 
Without loss of generality, assume that $x_i = 0$ and $y_i = 1$.

We now construct $x', y'$ by copying the highest bits up to (and including) bit 
$i$ from $x, y$, and clearing all remaining bits. By the properties of binary 
arithmetic, $y' = x' + 2^i$ and therefore $x' < y'$.

Assume by way of contradiction that we can now use the lower bits $0, \ldots, i - 1$
to change the ordering relation. Construct $x''$ by copying the highest bits up
to (and including) bit $i$ from $x'$, and setting all remaining bits. However,
$x'' = x' + 2^i - 1 = y' - 1$ and therefore $x'' < y'$ still holds.
\end{proof}

\begin{theorem}
All consecutive segments of size $2^\ell$, $0<\ell < k$, in $R_n^{rev}$
are order-equivalent.
\end{theorem}

\begin{proof}
Take any index $0 \leq s < n$, any $0 < \ell < k$, and any $0 \leq \off{i}, \off{j} < 2^\ell$ (all
arithmetic in this proof is modulo $n$).
Let indices $i = s + \off{i}$, $j = s + \off{j}$, $i' = i + 2^\ell$, and $j' = j + 2^\ell$, i.e.
we consider the segment $S$ of length $2^\ell$ starting at $s$ and its neighbor segment $S'$ starting
at $s' = s + 2^\ell$. $i, j$ are arbitrary nodes within $S$, and $i', j'$ their corresponding nodes
in $S'$.

We now prove that $S, S'$ are order-equivalent by showing that
$\rev{i} < \rev{j} \iff \rev{i'} < \rev{j'}$.

Notice that 

\begin{enumerate}[a)]
\item adding $2^\ell$ to $i$ ($j$) does not affect the $\ell$ least significant
      bits of $i$ ($j$) since $i$ ($j$) is an $n$-bit integer and addition is modulo $2^n$,
      and that \label{item:a}
\item in any sequence of $2^\ell$ consecutive numbers, the $\ell$ least significant
      bits are unique. \label{item:b}
\end{enumerate}

It follows from \ref{item:a} that the $\ell$ least significant bits of $i'$ ($j'$) are identical to those of
$i$ ($j$). These in turn are the $\ell$ most significant bits in $\rev{i}, \rev{j}, \rev{i'}, \rev{j'}$,
and it follows from \ref{item:b} and Lemma \ref{lemma:msb_order} that they 
completely determine the ordering relation of all elements in segments $S, S'$.

Since $S, S'$ are arbitrary consecutive segments, order-equivalence holds for
any, and thus all consecutive segments.
\end{proof}


% ------------------------------------------------------------------------------

\begin{Exc}{(A3.10.1.v1, 5 points):}
Consider an anonymous ring where processors start with binary
inputs. Prove rigorously that there is no uniform synchronous
algorithm for computing the AND of all the input bits.
\end{Exc}

\begin{proof}
By way of contradiction, assume that there exists an anonymous, uniform,
synchronous algorithm that computes the logical AND over the input of
all processors such that each participating processor knows the result upon termination,
and fix any such algorithm $A$.
Assume that it terminates with result $1$ in $r(n)$ rounds when executed on a ring $R$ consisting
of $n$ processors $p_0, \ldots, p_{n-1}$ with all inputs equalling $1$.

We now construct a ring $R'$ by combining a segment of $n + 2r(n)$ processors with input $1$
($\underline{p}_i, p_i, \overline{p}_i$) with a single processor with input $0$ ($q$):
$R' = 
(\underline{p}_0, \ldots, \underline{p}_{r(n) - 1},
 p_0, \ldots, p_{n-1},
 \overline{p}_0, \ldots, \overline{p}_{r(n) - 1},
 q)$.

Consider the first $r(n)$ rounds of an execution of $A$ on the ring $R'$. Since by anonymity
the $r(n)$-neighborhood
of each $p_i$ in $R'$ is equivalent to the $r(n)$-neighborhood for each $p_i$ in $R$, their state
after $r(n)$ rounds must be identical (Lemma 130 from the lecture slides),
i.e. they must have terminated with result $1$. This is however
a contradiction, since input $0$ of node $q$ forces the output of the algorithm to $0$.
\end{proof}

% ------------------------------------------------------------------------------

\begin{Exc}{(A3.10.3.v3, 7 points):}
Consider an anonymous ring where processors start with binary
inputs. Prove that $\Omega(n^2)$ is a lower bound for the
message complexity of any (non-uniform, message-driven)
asynchronous algorithm that computes the AND of all the
input bits. Show by means of a counter-example that this
lower bound does not hold for synchronous algorithms.
\end{Exc}

We will show the lower bound of $\Omega(n^2)$ messages for an anonymous, non-uniform,
asynchronous algorithm calculating the logical AND over all binary inputs by
constructing a highly symmetrical ring $S_n = (p_0, \ldots, p_{n-1})$ and showing
that any anonymous, non-uniform, synchronous algorithm $A$ which gains knowledge solely
through reception of messages (i.e. not through non-reception in a specific round)
must send at least $\Omega(n^2)$ messages. In the remaining exercise, we call
such algorithms ``ANSA algorithms''.
We assume that at the end of the algorithm, every processor must know the overall result.

\begin{definition}
The ring $S_n$ consists of $n$ processors $p_0, \ldots, p_{n-1}$ with inputs
$x_i = 1$, such that $p_{i+1}$ is the right neighbor
of $p_i$ for all $0 \leq i < n$ (index arithmetic is modulo $n$).
\end{definition}

\begin{definition}
Active rounds on $S_n$ are defined as in the book and lecture slides, i.e.
a round $r_k$ is active iff any processor sends a message in $r_k$.
$r_k$ is the $k$'th active round.
\end{definition}

\begin{lemma} \label{lemma:know_it_all}
In any ANSA algorithm $A$ on $S_n$, each processor must have knowledge of the 
input of every processor in the ring.
\end{lemma}

\begin{proof}
Fix any such algorithm $A$.
By contradiction, assume that some processor $p_i$ terminates with result $1$
while not knowing $x_{j \neq i}$. However, if $x_j = 0$, then $p_i$ would have to
terminate with result $0$, a contradiction.
\end{proof}

\begin{lemma} \label{lemma:knowledge_spread}
In any ANSA algorithm $A$ on $S_n$, in active round $r_k$ knowledge of any processor
$p_i$'s input $x_i$ has spread at most to processors at distance $k$ from
$p_i$.
\end{lemma}

\begin{proof}
Fix an arbitrary processor $p_i$.
By induction on distance $k$ from $p_i$.

The base case is $k = 0$, i.e. the initial state. In the initial state,
obviously only $p_i$ itself has knowledge of $x_i$.

Assume that in round $r_k$ knowledge of $x_i$ has spread at most to processors
distance $k$ from $p_i$. Since information is gained exclusively through messages,
a processor at distance $k$ must send a message to its neighbor at distance $k+1$
for it to receive knowledge of $x_i$. This can happen at the earliest in the
next active round $r_{k+1}$.
\end{proof}

\begin{lemma} \label{lemma:nr_rounds}
Any ANSA algorithm $A$ has at least $\lfloor \frac{n}{2} \rfloor$ active rounds.
\end{lemma}

\begin{proof}
Fix any such algorithm $A$ and an arbitrary processor $p_i$.

By Lemma \ref{lemma:know_it_all}, $p_i$ must know the input of every processor
in the ring. The maximal distance to any other processor in $S_n$ is
$\lfloor \frac{n}{2} \rfloor$. Let $p_j$ be such a processor at maximal distance
from $p_i$. By Lemma \ref{lemma:knowledge_spread}
it takes at least $\lfloor \frac{n}{2} \rfloor$ active rounds for $p_i$ to receive
knowledge of $x_j$.
\end{proof}

\begin{lemma} \label{lemma:identical_neighborhoods}
In any ANSA algorithm $A$ on $S_n$, the state of all $p_i, 0 \leq i < n$ is
identical in every round.
\end{lemma}

\begin{proof}
Fix any such $A$; since $A$ is anonymous and all processors have the same input,
they have identical $k$-neighborhoods for all $k$. By Lemma 130 from the lecture
slides, any pair of processors with identical $k$-neighborhoods are in the
same state in rounds $0, \ldots, r_k$.
\end{proof}

\begin{theorem} \label{theorem:and_lower_bound}
Any anonymous, non-uniform, asynchronous algorithm calculating the logical AND
must send $\Omega(n^2)$ messages.
\end{theorem}

\begin{proof}
Fix any ANSA algorithm $A$.

By Lemma \ref{lemma:nr_rounds}, $A$ must have at least $\lfloor \frac{n}{2} \rfloor$
active rounds. Since in each of these active rounds $r_k$ the state of all
$p_i \in S_n$ is identical (Lemma \ref{lemma:identical_neighborhoods}),
each $p_i$ sends a message in every active around, resulting in at least
$n \cdot \lfloor \frac{n}{2} \rfloor = \Omega(n^2)$ messages.

Lock-step synchronous execution
is a special case of asynchronous execution, hence the lower bound holds for
any asynchronous algorithm.
\end{proof}


\begin{theorem}
The lower bound of $\Omega(n^2)$ messages for an anonymous, non-uniform algorithm calculating the
logical AND of all binary inputs does not hold in synchronous systems.
\end{theorem}

\begin{proof}
The following algorithm $A$ for a ring of size $n$ with processors $p_i, 0 \leq i < n$
is given as a counterexample. 
Initially, all $p_i$ with input $0$ send $msg_0$ to their
left neighbor.
In rounds $1, \ldots, n-1$, processors with input $1$ simply forward all received
messages, while $p_i$ with input $0$ swallow all messages. In round $n$,
each processor terminates with result ``false'' if it has received a message, and
result ``true'' otherwise.

This algorithm calculates the logical AND with $O(n)$ message complexity since
each $p_i$ sends at most one message.
\end{proof}

% ------------------------------------------------------------------------------

\begin{Exc}{(S2.3.v0, 10 points):}
Consider the simple asynchronous leader election algorithm of Sec.~3.3.1 of
the textbook (Slide 104 of the lecture slides). Provide the complete
pseudo-code description and the transition relation of this algorithm,
and a complete and rigorous correctness proof (safety and liveness).
\end{Exc}

\begin{algorithm}
\caption{Simple asynchronous leader election} \label{alg:ale}
\begin{algorithmic}[1]
\Statex Code for processor $p_i, 0 \leq i < n$
\Statex Initially, $leader_i = $ null, and $id_i$ has been sent to the left neighbor
\For{each received message}
    \If{$leader_i \neq $ null}
        \State \textbf{continue}
    \EndIf

    \If{received message $id_j > id_i$}
        \State forward $id_j$ to left neighbor
    \ElsIf{received message $id_j = id_i$}
        \State $leader_i \gets id_i$
        \State send termination message with $id_i$ to left neighbor
    \ElsIf{received message $id_j < id_i$}
        \State do nothing, i.e. swallow message
    \ElsIf{received termination message with $id_j$}
        \State $leader_i \gets id_j$
        \State send termination message with $id_j$ to left neighbor
    \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

In this example, $id_i$ denotes the processor ID of $p_i$ for all $i$, and all
index arithmetic is modulo $n$.
The transition relations are as follows:

Variables $\in L_i$ of processor $p_i$:
\begin{itemize}
\item $leader_i$: a node ID or null
\end{itemize}

Initial state for all $i$:
\begin{itemize}
\item $leader_i = null$
\item $\inbuf_i[\cdot] = \emptyset$
\item $\outbuf_i[left] = id_i$
\item $\outbuf_i[right] = \emptyset$
\end{itemize}

Elementary transitions for each $p_i$:
\begin{itemize}
\item $\forall q_i \in Q_i: id_j \in \inbuf_i[right]: id_j > id_i$ and $leader_i = null$.
      Then $(q_i, \phi_i, q'_i) \in \Phi_i$,
      where $q'_i = q_i$ except:
      \begin{itemize}
      \item $\outbuf'_i[left] = \outbuf_i[left] \cup \{id_j\}$
      \item $\inbuf'_i[right] = \inbuf_i[right] \setminus \{id_j\}$
      \end{itemize}
\item $\forall q_i \in Q_i: id_j \in \inbuf_i[right]: id_j = id_i$ and $leader_i = null$.
      Then $(q_i, \phi_i, q'_i) \in \Phi_i$,
      where $q'_i = q_i$ except:
      \begin{itemize}
      \item $leader'_i = id_j$
      \item $\outbuf'_i[left] = \outbuf_i[left] \cup \{term_i\}$
      \item $\inbuf'_i[right] = \inbuf_i[right] \setminus \{id_j\}$
      \end{itemize}
\item $\forall q_i \in Q_i: id_j \in \inbuf_i[right]: id_j < id_i$ and $leader_i = null$.
      Then $(q_i, \phi_i, q'_i) \in \Phi_i$,
      where $q'_i = q_i$ except:
      \begin{itemize}
      \item $\inbuf'_i[right] = \inbuf_i[right] \setminus \{id_j\}$
      \end{itemize}
\item $\forall q_i \in Q_i: term_j \in \inbuf_i[right]$ and $leader_i = null$.
      Then $(q_i, \phi_i, q'_i) \in \Phi_i$,
      where $q'_i = q_i$ except:
      \begin{itemize}
      \item $leader'_i = id_j$
      \item $\outbuf'_i[left] = \outbuf_i[left] \cup \{term_j\}$
      \item $\inbuf'_i[right] = \inbuf_i[right] \setminus \{term_j\}$
      \end{itemize}
\item For all other $q_i \in Q_i$.
      Then $(q_i, \phi_i, q'_i) \in \Phi_i$,
      where $q'_i = q_i$ except:
      \begin{itemize}
      \item $\inbuf'_i[\cdot] = \emptyset$
      \end{itemize}
\end{itemize}

The set of compound transitions $\Phi'_i$ consists of the transitive closure over all elementary
transitions such that only a single compound transition is included for a given configuration.

Safety in the context of leader election algorithms means that at most one processor is in the
elected state at any time. This particular algorithm elects the processor with the maximal ID.
Let $p_0$ be that processor in ring $R = (p_0, \ldots, p_{n-1})$. Let $\outbuf_i$ be a shorthand for
$\outbuf_i[\text{left}]$.

\begin{theorem}
No processor other than $p_0$ can ever enter an elected state.
\end{theorem}

\begin{proof}
Observe that in order to enter an elected state, a processor $p_i$ must receive its own message
$id_i$, and that (since the message originated at $p_i$ itself and messages are sent only to the left)
this message must have traversed the entire ring. It follows from the code that the processor with
the maximal id $p_0$ will never forward a message $id_i < id_0$; therefore no $p_i, i \neq 0$ can
enter an elected state.
\end{proof}

Liveness of a leader election algorithm implies that eventually all processors terminate, and
at least one processor enters an elected state.

\begin{lemma} \label{lemma:le-liveness}
The message containing $id_0$ is eventually sent by each processor $p_i$ to $p_{i+1}$,
$0 \leq i < n$.
\end{lemma}

\begin{proof}
By induction on the processor position $i$.

In the initial state, $p_0$ has sent $id_0$ to $p_1$.

Assume the statement holds
for processor $p_{i-1}$; then for $0 < i < n$:

Since the execution is admissible, $id_0$ is eventually
delivered and processed at $p_{i}$. Since $id_0$ is maximal over all IDs, processor IDs
are unique, and $i > 0$, it must be forwarded by $p_i$ and is therefore sent to $p_{i+1}$.

It remains to show that no processor $p_{i \neq 0}$ terminates before it has 
forwarded $id_0$; by the code, such processors terminate only on receiving
the termination message. However, this message is only generated by $p_0$
once it enters an elected state, i.e. when $id_0$ has traversed the entire ring.
\end{proof}

\begin{theorem} \label{theorem:liveness1}
Processor $p_0$ eventually enters an elected state.
\end{theorem}

\begin{proof}
By lemma \ref{lemma:le-liveness}, $p_{n-1}$ eventually sends $id_0$ to $p_0$.
$p_0$ therefore receives its own id, sends a termination message to $p_1$,
and terminates as the leader.
\end{proof}

\begin{theorem}
Eventually all other processors $p_i, i > 0$ terminate.
\end{theorem}

\begin{proof}
By Theorem \ref{theorem:liveness1}, $p_0$ eventually enters the elected state,
upon which a termination message to $p_1$ is sent. Each remaining $p_i, i > 0$ terminates
as non-leader on receiving the termination message, and forwards it to the left neighbor.
By a similar induction proof as Lemma \ref{lemma:le-liveness}, the termination message
is eventually sent to all processors.
\end{proof}

\section{Review Feedback}

Thanks to both reviewers for their constructive comments.

\subsection{Review 1}.

This review caught the most major mistakes (leader election vs. AND, anonymity,
asynchronous algorithms don't have rounds), but missed details in exercise
4's transition relations and pseudo-code as well as the incorrect argument of exercise 3.
In general, it is somewhat lacking in detail;
for instance, it would be helpful to state why you found a certain section
confusing.

A part of exercise 1 was split into a separate lemma + proof. I also added a 
clarification about addition modulo $2^n$ in $n$-bit integers. I did not change
the proof in exercise 2 since I believe the structure is perfectly clear.

I don't agree that a proof is incomplete if intermediate steps in the sum
calculation are not included - these are very trivial to verify. In any case,
the proof has been rewritten since I found out that my initial idea
just doesn't work. The counterexample has also been corrected.

Again, I disagree that Theorem 3 would have required an induction proof. In any case,
much of this exercise (including Theorem 3) has been replaced by a corrected
version that does not rely on rounds.

Grade: 3

\subsection{Review 2}.

A helpful review that caught most minor and major issues, but missed the incorrect 
argument of exercise 3. I extracted a train of
thought in exercise 1 into Lemma 1, hopefully it should be clearer this way.

All changes proposed for exercise 2 were integrated.

In exercise 3, the counter-example
now gives an anonymous algorithm calculating the logical AND over all inputs,
and the main proof has been rewritten.

All changes proposed for exercise 4 were integrated.

Grade: 1-
